\documentclass[MS]{wfuthesis}
\usepackage{tikz} % The GOAT
\usetikzlibrary{calc}
\usepackage{titlesec}
\usepackage[utf8]{inputenc}
\usepackage{graphicx} % Required for inserting images
\usepackage{subcaption} % Allows for the creation of subfigures
\usepackage{standalone} % Allows including standalone .tex files
\usepackage{multicol} % Multiple columns of text/figures
\usepackage{amsmath, amssymb} % Math packages
\usepackage{array}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}
\usepackage{xcolor} % Allows the creation of custom colors
\usepackage{algorithm, algpseudocode} % To write algorithm pseudo code
\usepackage{etoolbox}
\usepackage[intoc]{nomencl} % To add abbreviations page, intoc adds it to the table of contents
\usepackage{etoolbox} % goes with the above package, to add groupings to the abbreviations
\usepackage[backend=biber, natbib=true, style=numeric, sorting=none]{biblatex} % Bibliography and bibliography adjacent texts
\usepackage{xpatch}
\usepackage{nicefrac}
\usepackage{stmaryrd} 
\usepackage{pgfmath}
\usepackage{scalerel}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{xinttools}
\usepgfplotslibrary{groupplots}


\title{\textbf{Advances in Tensor Decompositions: Fast Matrix Multiplication Algorithms And Parallel Adaptive Compression Techniques}}
\author{\textbf{João Victor de Oliveira Pinheiro}}
\date{May 2025}
\linespread{1.5}

\department{Computer Science}
\advisor{Grey Ballard, Ph.D.}
\chairperson{Aditya Devarakonda, Ph.D.}
\member{Frank Moore, Ph.D.}
\member{Ramakrishnan Kannan, Ph.D.}

\newcommand{\JP}[1]{\textcolor{purple}{\textbf{JP:} #1}}
\newcommand{\GB}[1]{\textcolor{red}{\textbf{GB:} #1}}
\newcommand{\AD}[1]{\textcolor{blue}{\textbf{AD:} #1}}
\newcommand{\FM}[1]{\textcolor{green}{\textbf{FM:} #1}}
\newcommand{\RK}[1]{\textcolor{orange}{\textbf{RK:} #1}}



%%%%%%%%%%%%%%%%%%%%%%%%%
%% My Own Shennenigans %%
%%%%%%%%%%%%%%%%%%%%%%%%%

\titleformat{\chapter}[display]
  {\bfseries\Large}  % Format for the chapter title
  {}                 % No "Chapter X" label
  {0pt}              % Space between number and title
  {\Large}           % Additional formatting for the title

\definecolor{mycustomgreen}{RGB}{44, 183, 66}
\definecolor{mycustomorange}{RGB}{255, 149, 1}
\definecolor{mycustompurple}{RGB}{181, 29, 227}

\makenomenclature
\renewcommand{\nomname}{\textbf{LIST OF ABREVIATIONS}}
\renewcommand\nomgroup[1]{%
  \item[\bfseries
  \ifstrequal{#1}{C}{CP Decompositions}{}%
  \ifstrequal{#1}{T}{Tucker Decompositions}{}%
]}

\algrenewcommand\alglinenumber[1]{{\sffamily\footnotesize#1}}

\makeatletter
\xpatchcmd{\algorithmic}{\itemsep\z@}{\itemsep=-1.5ex}{}{}
\makeatother

\addbibresource{references.bib}



%%%%%%%%%%%%
%% THESIS %%
%%%%%%%%%%%%

\begin{document}
    \include{setup}
    \maketitle
    %%%%%%%%%%%%%%%%%%%%%%
    %% Acknowledgements %%
    %%%%%%%%%%%%%%%%%%%%%%
    \section*{\textbf{ACKNOWLEDGEMENTS}} 
        Dedico este trabalho aos meus pais, que sob muito sol, fizeram-me chegar até aqui, na sombra.
        
        % To all of those that came before me, and to all of those who are yet to come

        % I would like to thank the Whitener family. Specifically the trouble child that is Nathan

        % I would like to thank Joe and Gabby not because they are physicis (boo)
        % but because they are kinda cool





    %%%%%%%%%%%%%%%%%%%%%%%
    %% Table of Contents %%
    %%%%%%%%%%%%%%%%%%%%%%%
    \tableofcontents
    \newpage





    %%%%%%%%%%%%%%%%%%%%%%%%%%%
    %% List of Illustrations %%
    %%%%%%%%%%%%%%%%%%%%%%%%%%%
    \listoffigures
    \newpage





    %%%%%%%%%%%%%%%%%%%%%%%%%%
    %% List of Abreviations %%
    %%%%%%%%%%%%%%%%%%%%%%%%%%
    \nomenclature[C]{CP:}{Cynical Polynomial}
    \nomenclature[T]{LLSV:}{Left Leading Singular Values}
    \nomenclature[T]{EVD:}{Eigenvalue Decomposition}
    \nomenclature[T]{SVD:}{Singular Value Decomposition}
    \nomenclature[T]{HOSVD:}{Higher Order Singular Value Decomposition}
    \nomenclature[T]{STHOSVD:}{Sequentially Truncated Higher Order Singular Value Decomposition}
    \nomenclature[T]{HOOI:}{Higher Order Orthogonal Iterations}
    \nomenclature[T]{HOSI:}{Higher Order Subspace Iterations}
    \nomenclature[T]{DT:}{Dimention Trees}

    \printnomenclature
    \newpage





    %%%%%%%%%%%%%%
    %% Abstract %%
    %%%%%%%%%%%%%%
    \addcontentsline{toc}{chapter}{\textbf{ABSTRACT}}
    \begin{center}
        \textbf{ABSTRACT}
    \end{center}
    Tensors are essential in modern-day computational and data sciences. This
    work presents recent advances in tensor decompositions which are techniques
    that break down complex high-dimensional arrays into smaller structured
    components. There are two projects presented in this thesis, each in its own
    chapter. The first project applies tensor decompositions on searches for new
    algorithms for fast matrix multiplication. The second project focuses on the
    development of scalable, adaptive methods for compressing massive data sets.
    Here, the abstracts for the two projects separately, in chronological order
    in which I began working on them.

    \textbf{Searching For Fast Matrix Multiply Algorithms using Cyclic-Invariant
    CP decomposition:} Fast matrix multiplication algorithms correspond to exact
    CP decompositions of tensors that encode matrix multiplication of fixed
    dimensions. This 3-way matrix multiplication tensor M has cyclic symmetry:
    the entry values are invariant under cyclic permutation of the indices. The
    CP decomposition of Strassen's original fast matrix multiplication algorithm
    for 2x2 matrices is cyclic invariant, and cyclic invariant decompositions
    are known to exist for 3x3 and 4x4 matrix multiplication as well. Cyclic
    invariance means a cyclic permutation of the CP factors results in the same
    CP components, just in a different order. We describe how to search for
    these solutions, which involve one third of the variables of generic
    solutions, using the damped Gauss-Newton optimization method along with
    heuristic rounding techniques, and we summarize the algorithms discovered so
    far.

    \textbf{Parallel Higher-Order Orthogonal Iteration for Tucker Decomposition
    with Rank Adaptivity:} Higher Order Orthogonal Iteration (HOOI) is an
    algorithm that uses block coordinate descent optimization to compress an
    input tensor into a Tucker format approximation. Classical HOOI requires
    specifying the ranks, or dimensions of the core tensor, and seeks to
    minimize the approximation error. 	We introduce HOOI-Adapt, a novel
    error-specified algorithm that adaptively selects the ranks of the Tucker
    tensor in order to maximize the compression subject to the error threshold.
    We implement HOOI-Adapt using the distributed-memory parallel TuckerMPI
    library and employ memoization to reduce the computational cost.
    Furthermore, we show that when the ranks are small, HOOI-Adapt has lower
    computational cost compared to the alternative Sequentially Truncated
    Higher-Order SVD (ST-HOSVD) algorithm. Our benchmarks demonstrate that our
    parallel implementation of HOOI-Adapt outperforms TuckerMPI's ST-HOSVD for
    small ranks when scaled to terabyte-sized input datasets on NERSC's
    Perlmutter platform.





    %%%%%%%%%%%%%%%%%%
    %% Introduction %%
    %%%%%%%%%%%%%%%%%%
    \chapters{}
    \chapter{\textbf{Introduction}}
        \input{text/chapter1/intro.tex}
        \newpage





    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    %% Chapter 2 - The CP Decomposition %%
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \chapter{\textbf{The CP Decomposition}}
        \input{text/chapter2/cp.tex}
        \newpage





    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    %% Chapter 3 - The Tucker Decomposition %%
    %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
    \chapter{\textbf{The Tucker Decomposition}}
        \input{text/chapter3/tucker.tex}
        \newpage





    %%%%%%%%%%%%%%%%
    %% References %%
    %%%%%%%%%%%%%%%%
    \addcontentsline{toc}{chapter}{\textbf{REFERENCES}}
    \printbibliography
    




    %%%%%%%%
    %% CV %%
    %%%%%%%%
    \addcontentsline{toc}{chapter}{\textbf{CURRICULUM VITAE}}

\end{document}
